Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-10 analysis/paper/paper.Rmd
edited the result section

The GC-MS analysis for organic residues extracted from Guijiabao potsherds revealed the presence of lipid compounds, but with low amounts in most of them. The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid ) and C18:0 (stearic acid), and monounsaturated fatty acid  C18:1 (oleic acid), are the main compounds for most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in most animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of C16:0 and C18:0 fatty acids in our samples suggest that the original organic residues were highly degraded during post-depositional processes. In addition, most potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A larger content of C18:1 and C16:0 with smaller content of C18:0 could indicate an origin of plant oils [@Mayyas2018; @Copley2005]. Although oleic acid (C18:1) has the highest concentration in most of our potsherd samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.

@@

Long-chain fatty acids from C20:0 to C32:0 are present in only one pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with an origin of degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate evidence of degraded animal fats or plant oil.

@@

@@

To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the δ^13^C values of C16:0 and C18:0 using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on different ways where dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the origins of degraded fats [@Evershed2008; @Salque2013]. The isotopic results show the Δ13C values in our samples range from `r delta_min` ‰ to `r delta_max` ‰ that were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

Most of the potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. There are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) that falls within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

```{r isotopic-plots, echo=FALSE, fig.cap= ""}
knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
```

@@

>>>>

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them. The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid) and C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds for most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of C16:0 and C18:0 fatty acids in our samples suggest that the original organic residues were highly degraded during post-depositional processes. In addition, some potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005]. Although oleic acid (C18:1) has the highest concentration in most of our potsherd samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.

@@

Long-chain fatty acids from C20:0 to C32:0 are present in only one pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

min_carbon_isotope <- round(min(delta_meth_cor$`13C C18:0`), 2)
max_carbon_isotope <- round(max(delta_meth_cor$`13C C16:0`), 2)

@@

To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on different ways where dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. It should be noted that there are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C16:0 and C18:0. B: The Δ13C (δ13C18:0-16:0) values plotted against δ13C16:0 values"}
knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
```

@@

1000 BC horse
Gaushan-90 pig, plant-most are rice: 95% and a few millet
